Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2024-02-05 Manuscript/Manuscript.tex
MS, version info

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  10pt,
  onecolumn]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Arial}
  \setmonofont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=0.8in,right=0.8in,top=0.8in,bottom=0.5in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\usepackage{svg}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

We identified a single \emph{nitric oxide synthase} (\emph{NOS}) gene in
the \emph{Platynereis dumerilii} genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that \emph{Platynereis}
NOS belongs to an orthology group of bilaterian NOS sequences (Figure
1---figure supplement 1). To characterise the expression pattern of
\emph{NOS}, we used in situ hybridization chain reaction (HCR) and
transient transgenesis. In two- and three-day-old larvae, we detected
\emph{NOS} expression in four cells (two of them weakly expressing) in
the apical organ region (Figure 1D and Figure 1---figure supplement 2).
\emph{NOS} was also expressed in the region of the visual eyes (adult
eyes) and the pigmented eyespots (Figure 1---figure supplement 2). The
four apical organ cells, but not the eyes, were also labelled with a
\emph{NOS} reporter construct driving palmitoylated tdTomato (Figure
1E). This reporter also revealed the axonal projections of these central
NOS-expressing neurons. The position and morphology of the four
\emph{NOS+} cells allowed us to identify the same four cells as four
interneurons (INNOS) in our three-day-old whole-body \emph{Platynereis}
volume EM data (Verasztó et al., 2020; Williams et al., 2017) (Figure
1B,C). In the synaptic connectome, the INNOS cells are postsynaptic to
the UV-sensory cPRCs and presynaptic to the INRGWa interneurons, which
are also cPRC targets (Figure 1C, F). The projections of INNOS cells are
segregated into input (dendritic) and output (axonal) compartments, with
cPRC inputs occurring on the dendritic part in the dense neurosecretory
plexus. INNOS output-synapses form in the more ventral projection region
(Figure 1---figure supplement 3). INRGWa neurons synapse on the head
serotonergic ciliomotor neurons (Ser-h1), which synapse on the
cholinergic MC ciliomotor neuron and cells of the prototroch ciliary
band (Figure 1C, F) (Verasztó et al., 2017).

\begin{figure}
\includegraphics[width=25in]{figures/Fig1} \caption{**Figure 1. Identification of *NOS*-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA; magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub; green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

The expression of \emph{NOS} in the INNOS interneurons in the cPRC
circuit suggests that NO signalling may be involved in
UV/violet-avoidance. To test this, first we asked whether NO is produced
during UV/violet stimulation of the larvae. We injected the fluorescent
NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while
exposing the region of cPRC cilia to 405 nm violet light. To mark cell
outlines, we coinjected mRNA encoding a red-fluorescent reporter
(RGECO), allowing the identification of the cPRCs. Following light
stimulation in the region of the ramified cilia of the cPRCs, we
detected an increase in DAF-FM fluorescence in the anterior
neurosecretory neuropil, the region of INNOS projections. This increase
did not occur in larvae in which we illuminated a control area (Figure
2).

\begin{figure}
\includegraphics[width=29.17in]{figures/Fig2} \caption{**Figure 2. NO produced by UV/violet stimulation to cPRCs.** (**A**) DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. The circles indicate the location of cPRC and control stimulation. The cPRCs are marked by thin lines. (**B**) DAF-FM fluorescence before and during 405 nm light stimulation. (**C**) Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs or a control area (ctr stim.). The purple box indicates the duration of 405 nm stimulation. Individual normalized traces (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation. <br> Figure 2 -- source data 1. DAF-FM fluorescence reads.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

We next tested whether NO signalling regulates UV/violet avoidance. To
achieve this, we generated two \emph{Platynereis} \emph{NOS} knockout
lines using the CRISPR/Cas9 method. We recovered two deletions
(NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early
stop codon and thus likely representing null alleles (Figure 3---figure
supplement 1A). We could establish a homozygous line for both mutations
indicating that NOS is not an essential gene in \emph{Platynereis}. To
quantify UV avoidance, we recorded the trajectories of freely swimming
wild type and mutant larvae in vertical columns, illuminated laterally
from two opposite sides with 395 nm UV light (Figure 3A and Figure
3---figure supplement 1B). As previously shown, wild-type larvae swim
downward following non-directional UV/violet light stimulation (Verasztó
et al., 2018). In contrast, both two- and three-day-old homozygous
\emph{NOS}-mutant larvae showed a strongly diminished UV-avoidance
response (Figure 3A, B and Figure 3---figure supplement 1B, C). This
phenotype is similar to the defective UV-avoidance of \emph{c-opsin1}
mutant larvae (Verasztó et al., 2018) and reveals a requirement for NOS
in UV-avoidance behaviour. Wild type but not mutant larvae also showed
an increase in swimming speed under UV light that may be due to downward
swimming trajectories (swimming in the direction of gravity) (Figure 3B
and Figure 3---figure supplement 1C). We also tested directional
phototaxis, by exposing larvae to 480 nm directional collimated light
from the top of the column. Three-day-old but not two-day-old
\emph{NOS}-mutant larvae also showed reduced phototactic behaviour,
suggesting a function for \emph{NOS} in the visual eyes that mediate
three-day-old phototaxis (Randel et al., 2014) (Figure 3D and Figure
3---figure supplement 1G).

To distinguish between an acute and developmental function of NOS in
light responses, we also tested larvae exposed to the NOS inhibitor
L-NAME. Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a
dose-dependent inhibition of UV avoidance. In contrast, phototaxis was
not affected (Figure 3C, E). Overall, our results indicate an acute
requirement for NOS signalling in UV-avoidance and a possible indirect,
developmental role in the visual system, reminiscent of the function of
NO signalling in \emph{Drosophila} eye development (Gibbs and Truman,
1998).

\begin{figure}
\includegraphics[width=33.33in]{figures/Fig3} \caption{**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.**  (**A**) Swimming trajectories of wild type (WT, n=32) and *NOS* mutant (*NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec after the onset of 395 nm stimulation from the side. (**B**) Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**C**) Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**D**) Vertical displacement in 30 sec bins of wild type and mutant (*NOSΔ11/Δ11* and *NOSΔ23/Δ23*) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. (**E**) Vertical displacement in 30 sec bins of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. <br> Figure 3 -- source data 1-5. Source data for panels A-E.}\label{fig:unnamed-chunk-3}
\end{figure}

@@

Next we aimed to identify the NO receptor in the cPRCs. NO generally
acts via soluble guanylate cyclases (sGC), belonging to the guanylate
cyclase family with a CYC domain (PFAM domain: PF00211). NO binding to
the heme group of sGC leads to increased cyclic guanosine monophosphate
(cGMP) production. Analysis of sGCs in \emph{Platynereis} indicated that
these genes are not expressed in any of the cells of the cPRC circuit
(not shown and (Verasztó et al., 2017)). Recently, Moroz and coworkers
reported an atypical but widely conserved family of guanylate cyclases
with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as
potential mediators of NO signalling (Moroz et al., 2020). To identify
NIT-GCs in \emph{Platynereis}, we searched transcriptome resources and
retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1
and 2). To analyse the relationship of these sequences to metazoan
NIT-GCs, we retrieved protein sequences with a CYC domain from the
transcriptome and genome databases of 45 metazoan and 2 choanoflagellate
species. We carried out cluster analysis and phylogenetic reconstruction
on a group of membrane-bound guanylate cyclases with sGCs as an
outgroup. In agreement with Moroz et al. (Moroz et al., 2020), we found
a group of GCs with NIT domains with representatives in placozoans,
cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans. The 15
\emph{Platynereis} sequences belonged to several deeply diverged clades
in the phylogenetic tree (Figure 4---figure supplement 1 and 2).

To characterise the expression of NIT-GCs, we used previously published
spatially-mapped single-cell transcriptome data (Achim et al., 2015;
Williams et al., 2017). Among the 15 NIT-GCs, two showed high and
specific expression in the cPRCs and one was expressed in the INNOS
cells (Figure 4---figure supplement 2). In the single-cell data, we
could identify the cPRCs by the specific expression of the markers
\emph{c-opsin1} and the \emph{pedal-peptide2 neuropeptide precursor}
(\emph{MLD proneuropeptide}) (Arendt et al., 2004; Williams et al.,
2017) (Figure 4---figure supplement 3A). The INNOS cells were identified
by NOS expression and spatial mapping in the brain (Achim et al., 2015).
We decided to focus on two NIT-GCs expressed in the cPRCs and with a
full-length sequence, \emph{NIT-GC1} and \emph{NIT-GC2}. To confirm the
single-cell data, we first carried out in situ hybridisation chain
reaction (HCR) with probes for \emph{NIT-GC1} and \emph{NIT-GC2} mRNA.
Both genes were specifically expressed in the four cPRCs, as confirmed
by co-labeling with an acetylated α-tubulin antibody and with an HCR
probe against \emph{pedal peptide 2/MLD proneuropeptide} (Figure 4C, D
and Figure 4---figure supplement 3A-C). To analyse the subcellular
localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and
affinity-purified polyclonal antibodies against a specific peptide
sequence from both proteins. In immunostainings, we found that NIT-GC1
was localise to the region corresponding to the axonal projections of
the cPRCs in the anterior neurosecretory plexus (Figure 4E).
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat
antibody raised against \emph{Platynereis} NOS revealed the localisation
of both proteins in dots in the neurosecretory neuropil (Figure
4---figure supplement 3D). In contrast, NIT-GC2 specifically labelled
the ramified sensory cilia of the cPRCs (Figure 4F). These different
subcellular localisations suggest that the two NIT-GCs are involved in
different intracellular signalling processes in the ciliary and axonal
regions of the cPRCs.

@@

To test the in vivo function of NIT-GC1 and NIT-GC2 in cPRC responses,
we combined Ca\textsuperscript{2+} imaging with morpholino-mediated
knockdowns. We used two translation-blocking morpholinos for each
\emph{NIT-GC} gene and tested knockdown efficiency by immunostaining
injected animals with the NIT-GC1 and NIT-GC2 antibodies (Figure
4---figure supplement 3E,F). For both genes, the morpholinos led to a
strong reduction in the respective antibody signal, confirming efficient
knockdown and antibody specificity.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  10pt,
  onecolumn]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Arial}
  \setmonofont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=0.8in,right=0.8in,top=0.8in,bottom=0.5in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

We identified a single \emph{nitric oxide synthase} (\emph{NOS}) gene in
the \emph{Platynereis dumerilii} genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that \emph{Platynereis}
NOS belongs to an orthology group of bilaterian NOS sequences (Figure
1---figure supplement 1). To characterise the expression pattern of
\emph{NOS}, we used in situ hybridization chain reaction (HCR) and
transient transgenesis. In two- and three-day-old larvae, we detected
\emph{NOS} expression in four cells (two of them weakly expressing) in
the apical organ region (Figure 1E and Figure 1---figure supplement 2A,
B). \emph{NOS} was also expressed in the region of the visual eyes
(adult eyes) and the pigmented eyespots (Figure 1---figure supplement
2A, B). The four apical organ cells, but not the eyes, were also
labelled with an upstream promoter sequence of the \emph{NOS} gene
construct driving membrane-targeted palmitoylated tdTomato reporter
(Figure 1F). This reporter also revealed the axonal projections of these
central NOS-expressing neurons. The position and morphology of the four
\emph{NOS+} cells allowed us to identify the same four cells as four
interneurons (INNOS) in our three-day-old whole-body \emph{Platynereis}
volume EM data (Verasztó et al., 2020; Williams et al., 2017) (Figure
1C, D). In the synaptic connectome, the INNOS cells are postsynaptic to
the UV-sensory cPRCs and presynaptic to the INRGWa interneurons
(cholinergic neurons that strongly express RGW neuropeptide), which are
also cPRC targets (Figure 1C, H). The projections of INNOS cells are
segregated into input (dendritic) and output (axonal) compartments, with
cPRC inputs occurring on the dendritic part in the dense neurosecretory
plexus (Figure 1D and Figure 1---figure supplement 3A). Immunostaining
revealed that the localization of NOS at the protein level is more
strongly concentrated in the dendrites of the input region even within
INNOS (Figure 1G and Figure 1---figure supplement 2C). INNOS
output-synapses form in the more ventral projection region (Figure 1D
and Figure 1---figure supplement 3). INRGWa neurons synapse on the head
serotonergic ciliomotor neurons (Ser-h1), which synapse on the
cholinergic MC ciliomotor neuron and cells of the prototroch ciliary
band (Figure 1C, H) (Verasztó et al., 2017).

\begin{figure}
\includegraphics[width=34.72in]{figures/Fig1} \caption{**Figure 1. Identification of *NOS*-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Volume rendering of the four INNOS neurons with incoming and outgoing synapses. **(E)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Expression of a *NOS* reporter (NOSp::palmi-3xHA; magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub; green) highlights cPRC cilia and the neuropil. **(G)** Immunostaining for NOS (magenta), co-stained for acetylated α-tubulin (acTub; green). **(H)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

The expression of \emph{NOS} in the INNOS interneurons in the cPRC
circuit suggests that NO signalling may be involved in
UV/violet-avoidance. To test this, first we asked whether NO is produced
during UV/violet stimulation of the larvae. We injected the fluorescent
NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while
exposing the region of cPRC cilia to 405 nm violet light (Figure
2---figure supplement 1). To mark cell outlines, we coinjected mRNA
encoding a red-fluorescent reporter (RGECO), allowing the identification
of the cPRCs. Following light stimulation in the region of the ramified
cilia of the cPRCs, we detected an increase in DAF-FM fluorescence in
the anterior neurosecretory neuropil, the region of INNOS projections.
This increase did not occur in larvae in which we illuminated a control
area (Figure 2).

\begin{figure}
\includegraphics[width=41.67in]{figures/Fig2} \caption{**Figure 2. NO produced by UV/violet stimulation to cPRCs.** (**A**) DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. The circles indicate the location of cPRC and control stimulation. The cPRCs are marked by thin lines. (**B**) DAF-FM fluorescence before and during 405 nm light stimulation. (**C**) Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs or a control area (ctr stim.). The purple box indicates the duration of 405 nm stimulation. Individual normalized traces (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation. <br> Figure 2 -- source data 1. DAF-FM fluorescence reads.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

We next tested whether NO signalling regulates UV/violet avoidance. To
achieve this, we generated two \emph{Platynereis} \emph{NOS} knockout
lines using the CRISPR/Cas9 method. We recovered two deletions
(NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early
stop codon and thus likely representing null alleles (Figure 3---figure
supplement 1A). We could establish a homozygous line for both mutations
indicating that NOS is not an essential gene in \emph{Platynereis}
(Figure 3---figure supplement 1B, C). To quantify UV avoidance, we
recorded the trajectories of freely swimming wild type and mutant larvae
in vertical columns, illuminated laterally from two opposite sides with
395 nm UV light (Figure 3A and Figure 3---figure supplement 2A, B). As
previously shown, wild-type larvae swim downward following
non-directional UV/violet light stimulation (Verasztó et al., 2018). In
contrast, both two- and three-day-old homozygous \emph{NOS}-mutant
larvae showed a significantly diminished UV-avoidance response (Figure
3B, C and Figure 3---figure supplement 1C, D). This phenotype is similar
to the defective UV-avoidance of \emph{c-opsin1} mutant larvae (Verasztó
et al., 2018) and reveals a requirement for NOS in UV-avoidance
behaviour. Wild type but not mutant larvae also showed a significantly
increase in swimming speed with an increase in ciliary beat frequency
under UV light that may be due to downward swimming trajectories
(swimming in the direction of gravity) (Figure 3---figure supplement 3).
We also tested directional phototaxis, by exposing larvae to 480 nm
directional collimated light from the top of the column (Figure
3---figure supplement 2A). Three-day-old but not two-day-old
\emph{NOS}-mutant larvae also showed reduced phototactic behaviour,
suggesting a function for \emph{NOS} in the visual eyes that mediate
three-day-old phototaxis (Randel et al., 2014) (Figure 3D and Figure
3---figure supplement 1E).

To distinguish between an acute and developmental function of NOS in
light responses, we also tested larvae exposed to the NOS inhibitor
L-NAME. Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a
dose-dependent inhibition of UV avoidance (Figure 3E, F). In contrast,
phototaxis was not affected (Figure 3G). Overall, our results indicate
an acute requirement for NOS signalling in UV-avoidance and a possible
indirect, developmental role in the visual system, reminiscent of the
function of NO signalling in \emph{Drosophila} eye development (Gibbs
and Truman, 1998).

\begin{figure}
\includegraphics[width=44.44in]{figures/Fig3} \caption{**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.**  (**A**) Swimming trajectories of wild type (WT, n=32) and *NOS* mutant (*NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec before the onset of 395 nm stimulation from the side. (**B**) Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**C**) Comparison of vertical position of wild type and mutant three-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. The wild type (WT, n=32) and *NOS* mutant (*NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) larvae. (**D**) Vertical displacement in 30 sec bins of wild type (10 batches) and mutant (*NOSΔ11/Δ11*, 5 batches and *NOSΔ23/Δ23*, 7 batches) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. (**E**) Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. (**F**) Comparison of vertical position of control (n=6) and L-NAME-treated (0.1 and 1 mM, n=6 each) three-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. (**G**) Vertical displacement in 30 sec bins of control (15 batches) and L-NAME-treated (0.1 and 1 mM, 15 batches each) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. <br> Figure 3 -- source data 1-7. Source data for panels A-G.}\label{fig:unnamed-chunk-3}
\end{figure}

@@

Next we aimed to identify the NO receptor in the cPRCs. NO generally
acts via soluble guanylate cyclases (sGC), belonging to the guanylate
cyclase family with a CYC domain (PFAM domain: PF00211). NO binding to
the heme group of sGC leads to increased cyclic guanosine monophosphate
(cGMP) production. Analysis of sGCs in \emph{Platynereis} showed that
these genes were not expressed in cPRC and INNOS in the cells of the
cPRC circuit, but only in INRGW (Verasztó et al., 2017). Recently, Moroz
and coworkers reported an atypical but widely conserved family of
guanylate cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376)
(NIT-GC) as potential mediators of NO signalling (Moroz et al., 2020).
To identify NIT-GCs in \emph{Platynereis}, we searched transcriptome
resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure
supplement 1 and 2). To analyse the relationship of these sequences to
metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from
the transcriptome and genome databases of 45 metazoan and 2
choanoflagellate species. We carried out cluster analysis and
phylogenetic reconstruction on a group of membrane-bound guanylate
cyclases with sGCs as an outgroup. In agreement with Moroz et al. (Moroz
et al., 2020), we found a group of GCs with NIT domains with
representatives in placozoans, cnidarians, some ecdysozoans,
echinoderms, and lophotrochozoans. The 15 \emph{Platynereis} sequences
belonged to several deeply diverged clades in the phylogenetic tree
(Figure 4---figure supplement 1 and 2).

To characterise the expression of NIT-GCs, we used previously published
spatially-mapped single-cell transcriptome data (Achim et al., 2015;
Williams et al., 2017). Among the 15 NIT-GCs, two showed high and
specific expression in the cPRCs and one was expressed in the INNOS
cells (Figure 4---figure supplement 2). In the single-cell data, we
could identify the cPRCs by the specific expression of the markers
\emph{c-opsin1} and the \emph{pedal-peptide2 neuropeptide precursor}
(\emph{MLD proneuropeptide}) (Arendt et al., 2004; Williams et al.,
2017) (Figure 4---figure supplement 3A). The INNOS cells were identified
by NOS expression and spatial mapping in the brain (Achim et al., 2015).
We decided to focus on two NIT-GCs expressed in the cPRCs and with a
full-length sequence, \emph{NIT-GC1} and \emph{NIT-GC2}. To confirm the
single-cell data, we first carried out in situ hybridisation chain
reaction (HCR) with probes for \emph{NIT-GC1} and \emph{NIT-GC2} mRNA.
Both genes were specifically expressed in the four cPRCs, as confirmed
by co-labeling with an acetylated α-tubulin antibody and with an HCR
probe against \emph{pedal peptide 2/MLD proneuropeptide} (Figure 4C, D
and Figure 4---figure supplement 3A-C). To analyse the subcellular
localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and
affinity-purified polyclonal antibodies against a specific peptide
sequence from both proteins. In immunostainings, we found that NIT-GC1
was localise to the region corresponding to the axonal projections of
the cPRCs in the anterior neurosecretory plexus (Figure 4E). In
contrast, NIT-GC2 specifically labelled the ramified sensory cilia of
the cPRCs (Figure 4F). These different subcellular localisations suggest
that the two NIT-GCs are involved in different intracellular signalling
processes in the ciliary and axonal regions of the cPRCs.

@@

To test the in vivo function of NIT-GC1 and NIT-GC2 in cPRC responses,
we combined Ca\textsuperscript{2+} imaging with morpholino-mediated
knockdowns. We used two translation-blocking morpholinos for each
\emph{NIT-GC} gene and tested knockdown efficiency by immunostaining
injected animals with the NIT-GC1 and NIT-GC2 antibodies (Figure
4---figure supplement 3D,E). For both genes, the morpholinos led to a
strong reduction in the respective antibody signal, confirming efficient
knockdown and antibody specificity (Figure 4---figure supplement 3F).

@@

\textbf{Ciliary beat frequency analysis}
